---
title: '**Using generalized additive models to analyze biomedical non-linear longitudinal data**'
subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
header-includes:
    \usepackage{placeins}
output:  
  bookdown::pdf_document2:
    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
    pandoc_args: --listings #calls the listings package to fit code within the page margins
    keep_tex: yes #keep LaTeX file for submission
    fig_caption: yes #allows captions in figures
    toc: false #do not include table of contents
    extra_dependencies:
      subfig: null #allows for subfigures
      breqn: null #line breaks for long equations
      caption: ["font={small}"] #size of the figure captions
      float: null #allows for control of placement of figures
    includes:
      in_header: preamble.sty #additional LaTeX formatting
  bookdown::word_document2:
    fig_caption: yes #figure caption
    keep_md: yes
  bookdown::html_document2:
    css: "style.css" #style for the HTML document
csl: elsevier-with-titles.csl #style for references
bibliography: refs.bib #references
link-citations: yes #adds links to the citations
'': default
---


# Abstract

In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear. In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We also present the basic theory of GAMs, and use simulated data that follows trends reported in the biomedical literature to demonstrate how these models are implemented in $\textsf{R}$ via the package _mgcv_, showing that GAMs are able to produce estimates that are consistent with the trends of non-linear data even if the case when missing observations exist. To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research.


```{r libraries, message = FALSE,include=FALSE}
#the packages needed to run the code

library(patchwork)
library(tidyverse)
library(mvnfast)
library(nlme)
library(mgcv)
library(gratia)
library(viridis)
set.seed(2021) #set seed for reproducibility

knitr::opts_chunk$set(fig.pos = 'H', out.extra = "",error=TRUE)

#global themes for plotting
thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting

#thm<-scale_color_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
#thm1<-scale_fill_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting

#this function will remove the default "author information" that is the Pandoc LaTeX template. Otherwise, and additional "and" will appear at the end of the author list.Or, the custom template that is in the YAML header and that has the "author" line deleted may be used as well. An explanation in detail is provided at:
#https://github.com/rstudio/bookdown/issues/1158
remove_author <- function(x) {
  # identify empty author line
  i <- grep("^\\\\author\\{\\}$", x)
  # be sure it is the one pandoc inserts
  if(length(i) != 0 && grepl('^\\\\date\\{', x[i+1])) x <- x[-i]
  x
}
options(bookdown.post.latex = remove_author)
```


```{r child='01-Background.Rmd'}
```

\FloatBarrier

```{r child='02-Challenges.Rmd'}
```

\FloatBarrier

```{r child='03-GAM Theory.Rmd'}
```

\FloatBarrier

```{r child='04-Longitudinal analysis with GAMs.Rmd'}
```

\FloatBarrier

```{r child='05-Discussion.Rmd'}
```

\FloatBarrier

```{r child='06-Conclusion.Rmd'}
```

\FloatBarrier

***


\newpage
# References

<div id="refs"></div>


```{r child='07-Appendix.Rmd'}

```






